Compression-based inference on graph data

### 2013-04-29 Benelearn/src/benelearn.tex
...

\usepackage{style/benelearn2013}
\usepackage{charter}
\usepackage{eulervm}
% \usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{subcaption}
\usepackage{float}
\usepackage{style/mlapa}
\usepackage{style/benelearn2013}

@@

\benelearntitlerunning{Compression-based inference on graph data}

@@

\begin{abstract}

We investigate the use of compression-based learning on graph data. Our hypothesis is that the sequential representation of data on which general purpose compressors operate, hampers compression-based learning, as a graph can be represented sequentially in many different, equivalent ways. Using Normalized Compression Distance (NCD), we test various compressors that operate directly on the graph representation against compressors that operate on a bitstring representation. We use both synthetic, randomly generated graphs and real-life datasets.

\end{abstract}

\emph{Compression equals learning} is the mantra of the field of compression-based learning. If we can compress data, we must have found some structure, we must have learned something. Conversely, if we have learned something about data, we should be able to use that knowledge to represent our data more succinctly.

@@

In this paper we investigate the use of these compression-based methods on graph data. Examples of such data are social graphs, transportation graphs, trade networks or semantic graphs. The graph is a powerful and versatile representation. Most applications of compression-based learning choose sequential models, such as deterministic finite automata or block-sorting compressors, which operate on bitstrings. If we have data in the form of a graph, we can translate it to a bitstring, but in this transformation we complicate our problem. For a different ordering of nodes, the graph is the same, but the bitstring changes radically. To a simple compressor, the two bitstrings may not share very much structure, even though they represent the same graph.

If this problem holds, we should investigate compressors that operate on the level of the graph representation, for instance by finding frequent subgraphs or clustering the graph. The methods do not suffer this problem of isomorphism, but as a result, they are more expensive than their sequential cousins. In this paper we hope to provide a first indication of how far the sequential approach will go, and whether the native approach will let us continue on from there.

@@

From the many methods born from Kolmogorov Complexity and Minimum Description Length, we choose \emph{Normalized Compression Distance} (NCD) as a representative method of compression-based learning. The choice is a practical one: NCD is a simple method which requires nothing more than a general purpose compressor. Any domain-specific knowledge we wish to use about our data (eg. it represents a graph) can be added to the compressor. We proceed under the assumption that conclusions reached about the performance of NCD on graph data will translate to MDL and other frameworks.

@@

The second principle is formalized in two ways. First, by `description' we mean a description in a formal language that is maximally expressive. To formalize expressivenes, we require that the method of description is Turing-complete (of equivalent strength to a Turing machine). By the (strong) Church-Turing thesis, this suggests that there is no reasonable way of defining a more expressive method of description.

@@

The result that makes Kolmogorov Complexity a useful measure of information content is that $K_U(\cdot)$ is only marginally dependent on the choice of $U$. If we suppose that there is another universal description method, $V$, we might ask what the expected difference is between $K_U(x)$ and $K_U(v)$. Let $k_V(x)$ be the shortest program for $x$ on $V$. Since, $U$ is universal, we know that is can compute $k_v$ by simulating $V$. This simulation is a program for $x$ on $U$ which is a bound for the shortest program on $U$:
\begin{align*}
K_U(x) 	&\leq |\langle T_v, k_v(x)\rangle|\\ 
		&=  |k_v(x)| + |v| + p(v) \\
		&=  K_V(x) + O(1)
\end{align*}

Where $p(v)$ is the penalty (ie. the additional bits) that the pairing function reuires to store its two arguments in a separable way. We require that this is only dependent on $v$. The final line shows that $K_U(\cdot)$ and $K_V(\cdot)$ differ only by a constant term, which is independent of $x$. To summarize: we may differ in opinion on how much information our data contains, but only by a constant amount.

@@

The length of the shortest program to get from $y$ to $x$ intuitively suggests that $K(\cdot\mid \cdot)$ can be seen as a similarity measure. Clearly, very little is required to transform a string into itself, or a very similar string, whereas for two random strings, only a program that stores the second in its entirety can make the transformation.

@@

The second issue is one of scale. If two strings of a million bits differ by 1000 bits, we might consider them quite similar, whereas two strings of 1000 bits that differ by that amount could not be more different. Thus, we would like to take the length of the strings into account. This gives us the Normalized information Distance (NID)

@@

Our aim is to test a sequential and graph-based compressor on a variety of graph data. To ascertain the performance of the compressors, we generate graphs from different sources, calculate their distances and see whether a clustering algorithm can reconstruct the original sources as clusters. Datasets and source code for these experiments are available. \footnote{\url{http://www.peterbloem.nl/benelearn2013}}

@@

We generate graphs from two models. The first is the classic Erd\H{o}s-R\'enyi (ER) model, where a uniform random choice is made from all graphs with $n$ nodes and $m$ links. The second is the Barab\'asi-Albert (BA) model \cite{albert2002statistical}, which grows a graph from a set of $n_0$ unconnected nodes, one node at a time, connecting each new node to $k$ distinct existing nodes where the probability that a given existing node is chosen for a connection is its degree, divided by the sum of the degrees of all nodes. Thus, under the BA model the more links a node has, the higher the probability that it will accrue even more. This effect causes the degree distribution of a BA network to form a scale free distribution.

@@

Once we have this dataset, consisting of four gold clusters, we calculate the NCD with a given compressor for every pair of graphs in the dataset, giving us a symmetric matrix. We use the k-medoids algorithm to cluster this set into two clusters, and we check whether graphs from a given source tend to end up in a specific cluster. As a random baseline, we generate a random distance matrix with every distance a uniform random value in $[0, 1)$, and run the clustering algorithm on that.

@@

We use the following datasets:
\begin{description}
	\item[cellular] The cellular network of the E. Coli bacterium. \cite{jeong2000large} Acquired from \url{http://www.nd.edu/~networks/resources/metabolic/index.html}
	\item[neural] The neural network of the C. Elegans nematode worm (ignoring link directions). \cite{achacoso1991ay, watts1998small} Acquired from \url{http://toreopsahl.com/datasets/#celegans}
	\item[co-purchase] A graph of items commonly purchased together on internet retailer Amazon.com. \cite{leskovec2007dynamics} Acquired from \url{http://snap.stanford.edu/data/amazon0302.html}
\end{description}

% Some basic properties of these graphs are given in table \ref{table:properties}.

%\begin{table}
%\label{table:properties}
%\begin{tabular}{ l | r r r r r}
%\hline
%    & nodes & links & C & $alpha$ & $p_\alpha$ \\
%\hline
%  web & & & & & \\
%  metabolic & & & & & \\
%  semantic A & & & & & \\
%  semantic B & & & & & \\
%  
%\hline
%\end{tabular}
%\caption{The basic properties of our natural datasets. We report the number of nodes, the number of links, the number of nodes in the largest component (C), the estimates power law exponent ($\alpha$), and the significance for the power law model ($p_\alpha$, significant for values \emph{above} 0.01).}

%\end{table}

@@

Subdue \cite{ketkar2005subdue} is an algorithm for finding frequent subgraphs in graph data. The algorithm searches through all possible subgraphs, searching for the  one that maximally compresses the data. The algorithm is essentially a beam search through the space of subgraphs.

Each subgraph serves as a model for compression by finding all occurrences of the subgraph in the data. Each occurrence is replaced by a single node, with the edges to the nodes containing annotations on how the edge shouldbe connected within the subgraph. We then store the subgrap once, together with the edited datagraph, to represent the original data. If the subgraph occurs often enough to offset the extra bits required to annotate the edges in the datagraph, the Subdue representation will be more efficient.

The following is a rough description of the algorithm:

\begin{pseudo}[h]
{
	$G$: the data graph \\
	$b$: the beam size \\
	\\
	$S \leftarrow [K_1]$ \textit{\#  initialize the list of substructures with a graph of a single node} \\
	\\
	\textbf{loop} \\
	\hspace*{5mm} $s \leftarrow $ \textit{head element of} $S$ \\
	\hspace*{5mm} \textit{add all extensions of} $s$ \textbf{to} $S$ \\

	\hspace*{5mm}\textit{sort $S$ by score($s', G$)} for $s' \in S$\\
	\hspace*{5mm}\textit{remove all but the first $b$ elements of $S$} \\
	\\
	\textbf{function score} $(s', G)$\\
	\hspace*{5mm} \textit{replace occurrences of $s'$ in $G$ with node $N$}\\
	\hspace*{5mm} \textit{annotate links to $N$ with the nodes in $s'$}\\
	\hspace*{5mm} \textbf{return} \textit{number of bits to store the edited $G$ and $s'$ }\\
}
	\caption{Pseudocode for the Subdue algorithm}
\end{pseudo}

The graph matching search (the first line of the \textbf{score} function) allows for inexact matches of the subgraph. In these cases, use a rough estimate of the number of bits required to transform the stored subgraph into the subgraph that is actually present in the data.

To make the calculation of a distance matrix feasible, we must find a way to reduce the computational costs of the Subdue algorithm. Our solution is to replace the exhaustive graph matching search by a non-exhaustive beam search. With this change, we cannot be sure that the score assigned to the given subgraph describes the optimal way to compress the data with that subgraph, but it does provide an upper bound.

@@

All graphs generated contain $100$ nodes. In the BA-model, we attach one node each step, giving $100$ links also. Note that this makes the BA graphs UAGs. For $k = 2$ the clustering problem would be more difficult. The fractal graphs we generate to depth $2$ by adding $4$ ancestors at each side of each link and $1$ extra link between the groups of ancestors. This results in networks of $90$ nodes and $100$ links.

@@

The Subdue algorithm has a lot of parameters. Wuring the search we return only one best subgraph. Our beamwidth at the top level is set to $5$. The beam width in the graph matching routine is set to $10$.

@@

\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l |  r r r r}
\hline
  ER                    &  0.083 & 0    & 0.17  & 0 \\
  BA                    &  0.16  & 0    & 0.083 & 0 \\
  fractal (pure)        &  0     & 0    & 0.083 & 0.17 \\
  fractal (small world) &  0     & 0.25 & 0     & 0 \\
\hline
\end{tabular}
\caption{Random baseline: error 0.46 (0.11)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l |  r r r r}
\hline
  ER                    &  0    & 0.25 & 0 & 0 \\
  BA                    &  0    & 0.25 & 0 & 0 \\
  fractal (pure)        &  0    & 0    & 0 & 0.25 \\
  fractal (small world) &  0.25 & 0    & 0 & 0 \\
\hline
\end{tabular}
\caption{GZIP: error 0.27 (0.12)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l |  r r r r}
\hline
  ER                    &  0    & 0.25 & 0    & 0    \\
  BA                    &  0    & 0    & 0.25 & 0    \\
  fractal (pure)        &  0    & 0    & 0    & 0.25 \\
  fractal (small world) &  0.25 & 0    & 0    & 0    \\
\hline
\end{tabular}
\caption{Subdue: error 0.14 (0.14)}
\end{subfigure}
\caption{Confusion matrices for vearious compressors. If we label the resulting clusters with so that the error is minimized (ie. reorder the columns of the confusion matrix to maximize the sum of the diagonal) we get an indication of the performance of the compressor captured in a single number. We report the mean over 10 experiments (and the standard deviation in brackets) below each confusion matrix. The confusion matrix is always for the first experiment.}

\label{table:synthetic-large}

\end{table}

\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l |  r r r}
\hline
  cellular      &  0.11 & 0.11 & 0.11 \\
  neural        &  0.11 & 0.11 & 0.11 \\
  co-purchase   &  0.22 & 0    & 0.11 \\
\hline
\end{tabular}
\caption{Random baseline: error 0.43 (0.11)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l |  r r r}
\hline
  cellular      &  0 & 0.33 & 0 \\
  neural        &  0.11 & 0.22 & 0 \\
  co-purchase   &  0.22 & 0 & 0.11 \\
\hline
\end{tabular}
\caption{GZIP: error 0.28 (0.17)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l | r r r}
\hline
  cellular      &  0    & 0.33 & 0 \\
  neural        &  0.33 & 0    & 0 \\
  co-purchase   &  0    & 0.33 & 0 \\
\hline
\end{tabular}
\caption{Subdue: error 0.34 (0.17) }
\end{subfigure}
\caption{Results for the experiment on natural datasets.}

\label{table:real-life}

\end{table}

% \subsection*{The models}

% One advantage of the model-based compressors over the general purpose ones is that we use the model as an indication of the structure in the data. In the case of graph data in particular, we can expect compressors that operate directly on the graph to produce models that are easier to relate to the semantics of the original data than those of the sequential compressors.

% To evaluate this we concatenate all graphs from a single source and report the model (ie. the frequent subgraph) found by Subdue.

@@

The notion of compression is a good framework within which to combine many approaches to inference from the most general to the most domain specific. The Minimum Description Length principle and its associated techniques, which have not been investigated yet for reasons of scope, offer the promise of an even broader field of approaches to the analysis of graph data.

\pagebreak

@@

>>>>

\usepackage[accepted]{style/benelearn2013}
%\usepackage{charter}
\usepackage{eulervm}
%\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{subcaption}
\usepackage{float}
\usepackage{style/mlapa}

@@

%\benelearntitlerunning{Compression}

@@

\begin{abstract}
We investigate the use of compression-based learning on graph data. General purpose compressors operate on bitstrings or other sequential representations. Since a single graph can be represented sequentially in many ways, we study the effect of this projection. Using Normalized Compression Distance (NCD), we test a sequential compressor versus a native graph compressor. We use both synthetic, randomly generated graphs and real-life datasets.  We conclude that, even under adverse circumstances, sequential representations contain enough structure for shallow algorithms to perform inference successfully. Algorithms that operate directly on the graph representation require a considerable increase in resources, but do allow for an increase in performance also.
\end{abstract}

The mantra of compression-based inference is that \emph{compression equals learning}. If we can compress data, we must have found some structure, we must have learned something. Conversely, if we have learned something about data, we should be able to use that knowledge to represent our data more succinctly. \footnote{Other features commonly associated with learning, such as generalization will, so the adherents of compression-based learning argue, be optimized along with the optimization of compression. See, for instance \cite{grunwald2005tutorial,grunwald2007minimum}.}

@@

In this paper we investigate the use of these compression-based methods on graph data. Examples of such data are social graphs, transportation graphs, trade networks or semantic graphs. The graph is a powerful and versatile representation. Most applications of compression-based learning use sequential models, such as deterministic finite automata or block-sorting compressors, which operate on bitstrings. If we have data in the form of a graph, we can translate it to a bitstring, of course, but in this transformation we complicate our problem. For different orderings of nodes, the graph is the same, but the bitstring changes radically. To a simple compressor, the two bitstrings may not share very much structure, even though they represent the same graph. We will call this the problem of isomorphism.

If this issue really `blinds' the sequential compressors to the structures in the graph, one option is to investigate compressors that operate on the level of the graph representation, for instance by finding frequent subgraphs or clustering the graph. The methods do not suffer this problem of isomorphism, but as a result, they are more expensive than their sequential cousins. In this paper we hope to provide a first indication of how far \emph{the sequential approach} will go, and whether \emph{the native approach} will let us continue on from there.

There is a wealth of research available on machine learning and data mining both within single graphs and on sets of graphs. See for instance, the book \cite{cook2006mining}. In this paper we do not share the data mining goal of extracting interesting features from the data, but only the goal of performing inference, using the compressors and their learning techniques as black boxes, and evaluating the the results of a chosen inference task.

@@

From the many machine learning methods based on principles of compression, we choose \emph{Normalized Compression Distance} (NCD) as a representative method of compression-based learning. The choice is a practical one: NCD is a simple method which requires nothing more than a general purpose compressor. Any domain-specific knowledge we wish to use about our data (eg. it represents a graph) can be added to the compressor. We proceed under the assumption that conclusions reached about the performance of NCD on graph data will translate to MDL and other frameworks.

@@

The second principle is formalized in two steps. First, by `description' we mean a description in a formal language that is maximally expressive. To formalize expressivenes, we require that the method of description is Turing-complete (of equivalent strength to a Turing machine). By the strong Church-Turing thesis, this suggests that there is no reasonable way of defining a more expressive method of description.

@@

The result that makes Kolmogorov Complexity a useful measure of information content is that $K_U(\cdot)$ is only marginally dependent on the choice of $U$. If we suppose that there is another universal description method, $V$, we might ask what the expected difference is between $K_U(x)$ and $K_U(v)$. Let $k_V(x)$ be the shortest program for $x$ on $V$. Since, $U$ is universal, we know that it can compute $k_v$ by simulating $V$. This simulation is a program for $x$ on $U$ which is a bound for the shortest program on $U$:
\begin{align*}
K_U(x) 	&\leq |\langle T_v, k_v(x)\rangle|\\ 
		&=  |k_v(x)| + |v| + p(v) \\
		&=  K_V(x) + O(1)
\end{align*}

Where $p(v)$ is the penalty (ie. the additional bits) that the pairing function requires to store its two arguments in a separable way. We require that this is only dependent on $v$. The final line shows that $K_U(\cdot)$ and $K_V(\cdot)$ differ only by a constant term, which is independent of $x$. To summarize: we may differ in opinion on how much information our data contains, but only by a constant amount.

@@

The length of the shortest program to get from $y$ to $x$ intuitively suggests that $K(\cdot\mid \cdot)$ can be seen as a similarity measure. Clearly, very little is required to transform a string into itself, or into a very similar string, whereas for two random strings, only a program that stores the second in its entirety can make the transformation.

@@

The second issue is one of scale. If two strings of a million bits differ by $1000$ bits, we might consider them quite similar, whereas two strings of $1000$ bits that differ by that amount could not be more different.\footnote{Note that this is only an intuitive example. If two strings differ in exactly every bit,a very short program transforms one into the other, so by NID, they are very similar.} Therefore, we would like to take the length of the strings into account. This gives us the Normalized information Distance (NID)

@@

Our aim is to test a sequential and a graph-based compressor on an inference tasks for a variety of graph data. To ascertain the performance of the compressors, we generate graphs from different sources, calculate their NCD distances and see whether a clustering algorithm can reconstruct the original sources as clusters. Datasets and source code for these experiments are available. \footnote{\url{http://www.peterbloem.nl/benelearn2013}}

\subsection*{Node ordering}

An important and subtle concern is the ordering of nodes in the sequential representation of our graphs. This issue is detailed very well in \cite{kang2011beyond}. As shown, there are various algorithms to determine node orderings that bring out a lot of the graph's inherent structure in the adjacency matrix, allowing a sequential compressor to use it. We can use substantial resources to find a good ordering of nodes to improve the performance of the sequential compressor, but if we did, we could no longer consider it a `shallow model'.

Since we are testing the capacity of the general purpose compressor to perform inference despite the problem of isomorphism, we will present it with a worst case scenario. We will use a random ordering of nodes for all graphs. If the general purpose compressor still outperforms the random baseline under these conditions, it will tell us that it is, at least in part, resistant to the problem of isomorphism.

@@

We generate graphs from four models.

The first is the classic Erd\H{o}s-R\'enyi (ER) model, where a uniform random choice is made from all graphs with $n$ nodes and $m$ links. The second is the Barab\'asi-Albert (BA) model \cite{albert2002statistical}, which grows a graph from a set of $n_0$ unconnected nodes, one node at a time, connecting each new node to $k$ distinct existing nodes where the probability that a given existing node is chosen for a connection is its degree, divided by the sum of the degrees of all nodes. Thus, under the BA model the more links a node has, the higher the probability that it will accrue even more. This effect causes the degree distribution of a BA network to form a scale free distribution.

@@

Once we have this dataset, consisting of four gold clusters, we calculate the NCD with a given compressor for every pair of graphs in the dataset, giving us a symmetric matrix. We use the k-medoids algorithm to cluster this set into four clusters.

To assess the performance of the clustering we label the clusters so that the accuracy is maximized (essentially assigning the optimal labels). Clearly, this would be cheating when testing a classifier, but since we are onnly interested in the clustering aspect, it gives us a straightforward performance measure.

As a random baseline, we generate a random distance matrix with every distance a uniform random value in $[0, 1)$, and run the clustering algorithm on that.

@@

We use the following datasets:
\begin{description}
	\item[cellular] The cellular network of the E. Coli bacterium. \cite{jeong2000large} Acquired from \url{http://www.nd.edu/~networks/resources/metabolic/index.html}
	\item[neural] The neural network of the C. Elegans nematode worm (ignoring link directions). \cite{achacoso1991ay,watts1998small} Acquired from \url{http://toreopsahl.com/datasets/#celegans}
	\item[co-purchase] A graph of items commonly purchased together on internet retailer Amazon.com. \cite{leskovec2007dynamics} Acquired from \url{http://snap.stanford.edu/data/amazon0302.html}
\end{description}

@@

Subdue \cite{ketkar2005subdue} is an algorithm for finding frequent subgraphs in graph data. The algorithm searches through all possible subgraphs, searching for the  one that maximally compresses the data. The algorithm is essentially a beam search through the space of subgraphs. If consists of three main routines:

\begin{description}
\item[Subgraph matching] This is essentially an algorithm for finding the occurrences of a given subgraph in a graph. The algorithm used is detailed in \cite{bunke1983inexact}. Since this is essentially a semi-exhaustive search for the solution to the NP-Hard problem of subgraph isomorphism, the matching can only be solved for very small subgraphs. Unfortunately, weven with subgraphs of four or five nodes, the matching is too slow in combination with the number of times it is executed to calculate a full distance matrix. To combat this issue, we remove all but the first $b_{\mbox{inner}}$ elements from the search queue after it is sorted at ech iteration, effectively turning the algorithm into a beam search.
\item[MDL Scoring]
This routine takes a subgraph, finds its occurrences in the data by the previous routine and replaces these occurrences with a single node with a new label, which we call a \emph{symbol node}. All links that connected to nodes in the subgraph are now connected to the symbol node. The links connecting to the symbol node are annotated with the indices of the nodes in the subgraph they were connected to. Storing the subgraph once together with the reduced data graph is enough to reconstruct the original data, and, if the subgraph occurs often, will lead to a significant compression.
\item[Subgraph search]
This is the `outer loop' of the algorithm. Starting with a graph of a single link beteen two nodes, it searches through the space of all connected graphs by extending each current candidate by one link at a time (possibly by adding a new node as well). The buffer of current candidates is sorted by MDL score, and the candidate with the highest score is extended to create new candidates. At each iteration all but the top $b_{\mbox{outer}}$ candidates are removed, turning the search into a beam search.
\end{description}

Algorithm \ref{pseudo:algorithm} shows a broad description of the whole algorithm.

\begin{pseudo}[h]
{
	$G$: the data graph \\
	$b_{\mbox{outer}}$: the beam size \\
	\\
	$S \leftarrow [K_1]$ \textit{\#  initialize the list of substructures with a graph of a single node} \\
	\\
	\textbf{loop} \\
	\hspace*{5mm} $s \leftarrow $ \textit{head element of} $S$ \\
	\hspace*{5mm} \textit{add all extensions of} $s$ \textbf{to} $S$ \\

	\hspace*{5mm}\textit{sort $S$ by score($s', G$)} for $s' \in S$\\
	\hspace*{5mm}\textit{remove all but the first $b$ elements of $S$} \\
	\\
	\textbf{function score} $(s', G)$\\
	\hspace*{5mm} \textit{replace occurrences of $s'$ in $G$ with node $N$}\\
	\hspace*{5mm} \textit{annotate links to $N$ with the nodes in $s'$}\\
	\hspace*{5mm} \textbf{return} \textit{\# of bits to store the edited $G$ and $s'$ }\\
}
	\caption{Pseudocode for the Subdue algorithm}

	\label{pseudo:algorithm}
\end{pseudo}

The graph matching search (the first line of the \textbf{score} function) allows for inexact matches of the subgraph. In these cases, we use a rough estimate of the number of bits required to transform the stored subgraph into the subgraph that is actually present in the data.

@@

All graphs generated contain $100$ nodes. In the BA-model, we attach one node each step, giving $100$ links also. Note that this makes the BA graphs UAGs. For $k = 2$ the clustering problem would be more difficult. The random graphs are generated with the exact same number of nodes and links. The fractal graphs we generate to depth $2$ by adding $4$ ancestors at each side of each link and $1$ extra link between the groups of ancestors. This results in networks of $90$ nodes and $100$ links.

@@

The Subdue algorithm has a lot of parameters. During the search we return only one best subgraph. Our beamwidth at the top level ($b_{\mbox{outer}}$) is set to $5$. The beam width in the graph matching routine ($b_{\mbox{inner}}$) is set to $10$.

@@

\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l |  l l l l}
\hline
  ER                    &  0.17  & 0.083   & 0     & 0 \\
  BA                    &  0.083 & 0.17    & 0     & 0 \\
  fractal (pure)        &  0.083 & 0       & 0.17  & 0 \\
  fractal (small world) &  0     & 0       & 0     & 0.25 \\
\hline
\end{tabular}
\caption{Random baseline: error 0.46 (0.11)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l |   l l l l}
\hline
  ER                    &  0    & 0.25 & 0 & 0 \\
  BA                    &  0    & 0.25 & 0 & 0 \\
  fractal (pure)        &  0    & 0    & 0 & 0.25 \\
  fractal (small world) &  0.25 & 0    & 0 & 0 \\
\hline
\end{tabular}
\caption{GZIP: error 0.27 (0.12)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l |  l l l l}
\hline
  ER                    &  0    & 0.25 & 0    & 0    \\
  BA                    &  0    & 0    & 0.25 & 0    \\
  fractal (pure)        &  0    & 0    & 0    & 0.25 \\
  fractal (small world) &  0.25 & 0    & 0    & 0    \\
\hline
\end{tabular}
\caption{Subdue: error 0.14 (0.14)}
\end{subfigure}
\caption{Confusion matrices for various compressors. Columns represent the clusters found by the k-medioids algorithm. We have labeled the resulting clusters so that the error is minimized (ie. reordered the columns of the confusion matrix to maximize the sum of the diagonal). We report the mean error (1 - the sum of the diagonal) over 10 experiments (and the standard deviation in brackets) below each confusion matrix. The confusion matrix shown is always for the first experiment in the run.}
\label{table:synthetic-large}
\end{table}

\begin{table}

\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l | l l l }
\hline
  cellular      &  0.11 & 0.11 & 0.11 \\
  neural        &  0.11 & 0.11 & 0.11 \\
  co-purchase   &  0.22 & 0    & 0.11 \\
\hline
\end{tabular}
\caption{Random baseline: error 0.43 (0.11)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l | l l l }
\hline
  cellular      &  0 & 0.33 & 0 \\
  neural        &  0.11 & 0.22 & 0 \\
  co-purchase   &  0.22 & 0 & 0.11 \\
\hline
\end{tabular}
\caption{GZIP: error 0.28 (0.17)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l | l l l }
\hline
  cellular      &  0    & 0.33 & 0 \\
  neural        &  0.33 & 0    & 0 \\
  co-purchase   &  0    & 0.33 & 0 \\
\hline
\end{tabular}
\caption{Subdue: error 0.34 (0.17) }
\end{subfigure}
\caption{Results for the experiment on natural datasets.}
\label{table:real-life}
\end{table}

@@

The notion of compression-based learning is a good framework within which to combine many approaches to inference from the most general to the most domain specific. The Minimum Description Length principle and its associated techniques, which have not been investigated yet for reasons of scope, offer the promise of an even broader field of approaches to the analysis of graph data.

@@

We would like to thank the reviewers for many insightful comments.

\appendix

\section*{Appendix: Graph coding}

The method of coding data is always a sensitive point in compression-based learning. The precise choices made in translating the data to a string of bits can greatly affect which patterns are picked up, or ignored by the subsequent inference procedure. Here we detail the exact procedure used to encode the graphs in both compressors.

\subsection*{GZIP}
We take half of the adjacency matrix ($(n^2 + n)/2$ bits), and store it in flattened form as an array of java byte primitives, together with the size of the string (since the stored representation is padded to a multiple of eight). We then serialize this representation into a java GZIPOutputStream. To make our implementation generic, we assume that the nodes and links are labeled, and serialize the labels in a fixed order after the adjacency matrix. In the graphs mentioned in this paper, there is a single label assigned to all elements, so inference shouldn;t be influenced.

\subsection*{Subdue}

In the subdue case we do our coding in a more precise way, without relying on platform-specific functions. More importantly, we only count the bits required to store our graph, rather than actually constructing the representation itself.

\subsubsection*{Plain graph}

To store a plain graph, we follow roughly the coding strategy outlined in \cite{holder1994substructure}.

We first store the number of nodes, and the maximum number of neighbours for a node in the graph $n_{\mbox{max}}$. We then store the adjacency matrix row by row.

For node $i$, we only need to store the connections to the $i$ nodes below it including itself. We use $l(n_{\mbox{max}})$ bits to store the number of such neighbours $n_i$, and $\log{i \choose n_i}$ bits to store the configuration of those neighbours.

We then account for the possibility of multiple links between node pairs. Since we expect this to be unlikely, we want to ensure that we do not waste bits coding for this occasion if it does not occur. We store the pairs for which there are multiple links as a list consisting of two numbers per element: the first is the index of the pair in a canonical ordering, the second is the number of extra links that exist for that pair (in prefix-coding). We use a single bit to indicate whether this list will follow, and we reserver a codeword to signal the end of the list.

After this, all that remains is to encode the labels of the nodes and links. We assume that the sender and receiver in our coding scheme posess a codebook that is efficient for the data given, so that if a label $l$ occurs with frequency $\#l$, we can encode it in $-\log \frac{\#l}{\sum_k \#k}$ bits.

We assume that there is some canonical ordering among the nodes and links, and store them as a stream. Since the number of labels is known from the adjacency matrix, this code is self-delimiting.

\subsubsection*{Graph with substructure}

To store the graph with a substructure, we first store the substructure itself. Since this contains no special symbols, we can store it simply using the above method (except we use the codebook based on the whole graph rather than the substructure to encode the labels). This is a prefix code, so we can start encoding the rest of the graph right after.

In the rest of the graph, we replace the occurrences of the substructure with a symbol node and store this graph by the method detailed for plain graphs. We store the number of symbol nodes in prefix coding, and only encode labels for the rest of the nodes (again, by the codebook of the full data).

Finally, we must store, for every link leading into a symbol node, which node in the substructure it should connect to to restore the original data. Since we know the graph and all the symbol nodes, we can presume there to be a canonical ordering in these links, and we use $\log n_s$ bits for each, where $n_s$ is the size of the substructure in nodes.
